---
title: "Transact SQL (T-SQL) "
subtitle: "Quick Reference Guide"
output:
  html_document:
    keep_md: false
    df_print: paged
    css: www/style.css
    toc: true
    toc_float: 
          collapsed: false
          smooth_scroll: false
    number_sections: no
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(IRdisplay)
library(htmltools)
```

```{r warning=FALSE, echo=FALSE, message=FALSE}

title_and_header_dir <- "www/title_and_header.html"

title_and_header <- readLines(title_and_header_dir, warn = FALSE)
HTML(paste(title_and_header, collapse = "\n"))
```

```{r message=FALSE, warning=FALSE, include=FALSE}

# Establish the connection with SQL Server
library(RODBC)
library(DBI)
library(odbc)

server = "JOAONETO\\SQL_JNETO" # SQL Server name
database = "corporation" # database name
table = "employees" #table name

conn <- dbConnect(
  odbc::odbc(), 
  driver = "SQL Server",
  server = server,
  database = database,
  #uid = "your_username",
  #pwd = "your_password
  Trusted_Connection = "yes"
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# Close the database connection
#dbDisconnect(conn = conn)
```

## 1. Database Management

### Create tables

```{sql, connection = conn}
/*
CREATE TABLE [omni_company].[dbo].[friends] (
    id INT NOT NULL,
    first_name VARCHAR(30),
    last_name VARCHAR(30),
    birthdate DATE
)
*/
```

### Insert rows

```{sql, connection = conn}
/*
INSERT INTO friends VALUES(1, 'Irina', 'Munroe', '1949-05-30')
INSERT INTO friends VALUES(2, 'Joel', 'Martin', '1945-04-30')
INSERT INTO friends VALUES(3, 'Jonh', 'Smith', '1940-05-30')
*/
```

```{sql, connection = conn, echo = FALSE, message=FALSE, warning=FALSE, include=TRUE}
/*
SELECT *
FROM [omni_company].[dbo].[friends] 
*/
```

```{sql, connection = conn}
--DROP TABLE [omni_company].[dbo].[friends] 
```

## 2. Database Querying & Analysis

### Filter & Sorting

```{sql, connection = conn}
SELECT *
FROM corporation.dbo.employees
```

### Date datatype manage

The '***DATEPART()***' function returns an integer value, extracting the specified date part of the specified date.

The '***GETDATE()***' function returns the current database system date and time (YYYY-MM-DD hh:mm:ss.mmm).

```{sql, connection = conn}
SELECT 
  first_name, 
  last_name, 
  start_date,
  DATEPART(year, start_date) 'year',
  DATEPART(month, start_date) 'month',
  DATEPART(day, start_date) 'day',
  DATEPART(hour, GETDATE()) 'hour',
  DATEPART(minute, GETDATE()) 'minute',
  DATEPART(second, GETDATE()) 'second'
FROM corporation.dbo.employees
```


### Create variables

How many employees started in March of 2001?
```{sql, connection = conn, warning = FALSE, message= FALSE}
DECLARE @year INT, @month INT;
SET @year = 2001
SET @month = 4

SELECT first_name, last_name, department, start_date
FROM corporation.dbo.employees
WHERE 
  DATEPART(year, start_date) = @year AND 
  DATEPART(month, start_date) = @month
```


### Grouping

### Joins

### CTE (Common Table Expression)

### Window functions

Show for each employee their salary together with the minimum and 
maximum salaries of employees in their department.
```{sql, connection = conn, warning = FALSE, message= FALSE}
SELECT 
	first_name, 
	last_name,
	department,
	salary,
	MIN(salary) OVER (PARTITION BY department) AS min_salary_dep,
	MAX(salary) OVER (PARTITION BY department) AS max_salary_dep
FROM corporation.dbo.employees
ORDER BY min_salary_dep
```


Show for each employee the number of employees who are members of the 
same department as them.
```{sql, connection = conn, warning = FALSE, message= FALSE}
SELECT 
	first_name,
	last_name, 
	department,
	COUNT(id) OVER (PARTITION BY department) 'num_employees_dep'
FROM corporation.dbo.employees
ORDER BY num_employees_dep DESC
```

Show for each employee the number of employees who started in the same 
month as them.
```{sql, connection = conn, warning = FALSE, message= FALSE}
SELECT 
	first_name, 
	last_name,
	start_date,
	COUNT(id) OVER(
	PARTITION BY DATEPART(month, start_date), DATEPART(year, start_date)) 'num_employees'
FROM corporation.dbo.employees
WHERE DATEPART(year, start_date) >= 1990
```

Get a table of employees showing the order in which they started work with the corporation split by department 
```{sql, connection = conn, warning = FALSE, message= FALSE}
SELECT 
	first_name,
	last_name, 
	department,
	start_date,
	RANK() OVER (PARTITION BY department ORDER BY start_date ASC) 'rank_department'
FROM corporation.dbo.employees
WHERE DATEPART(year, start_date) >= 1990
```


Get a table of employee id, first and last name, grade and salary, together with two new columns showing the maximum salary for employees of their grade, and the minimum salary for employees of their grade.
```{sql, connection = conn, warning = FALSE, message= FALSE}
SELECT 
	id,
	first_name,
	last_name,
	grade, 
	salary,
	MAX(salary) OVER (PARTITION BY grade) 'max_salary_grade',
	MIN(salary) OVER (PARTITION BY grade) 'min_salary_grade'
FROM employees
WHERE grade IS NOT NULL
ORDER BY id ASC
```


Add a column for each employee showing the ratio of their salary to the average salary of their team.
```{sql, connection = conn, warning = FALSE, message= FALSE}

SELECT 
	e.first_name,
	e.last_name,
	e.salary,
	t.name,
	AVG(e.salary) OVER (PARTITION BY t.name) AS avg_salary_team,
	e.salary/AVG(e.salary) OVER (PARTITION BY t.name) 'ratio'
	--CONVERT(numeric(10,2), e.salary/AVG(e.salary) OVER (PARTITION BY t.name)) AS avg
FROM employees AS e LEFT JOIN teams AS t
ON e.team_id = t.id	
WHERE e.salary IS NOT NULL
```